Short-time simulations of open collective quantum spin systems

### 2019-04-18 soc_squeezing/collective_paper/collective_spin.tex
added citation to squeezing paper on arxiv; added tutorial for computing a correlator

\newcommand{\A}{\mathcal{A}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\S}{\mathcal{S}}

\newcommand{\N}{\mathbb{N}}

@@

In this work, we present a new method to compute short-time dynamics
of collective spin systems with generic Markovian decoherence.  The
only restriction on decoherence is that, like the coherent collective
dynamics, it must act identically on all constituent particles.  This
method is based on a short-time expansion of exact solutions to the
equations of motion for Heisenberg operators.  Evaluating this
expansion requires knowing the structure constants of a collective
spin operator algebra; the calculation of these structure constants
(in Appendices \ref{sec:identities}--\ref{sec:prod_general}) is one of
the main technical results of this work, which we hope will empower
both analytical and numerical studies of collective spin systems in
the future.  We benchmark our method against both exact and quantum
trajectory Monte Carlo computations of spin squeezing in accessible
parameter regimes, highlighting both advantages and limitations of our
short-time expansion.  Finally, we showcase applications of our method
with computations that are inaccessible to other numerical methods.

@@

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$\hat s_{\a=\x,\y,\z}\equiv\hat\sigma_\a/2$ and
$\hat s_\pm\equiv\hat s_\x\pm i\hat s_\y=\hat\sigma_\pm$ with Pauli
operators $\hat\sigma_\a$, we denote an operator that acts with
$\hat s_\a$ on the spin indexed by $j$ and trivially (i.e.~with the
identity $\hat\1$) on all other spins by $\hat s_\a^{(j)}$.  We then
define the collective spin operators
$\hat S_\a\equiv\sum_{j=1}^N\hat s_\a^{(j)}$ for
$\a\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\hat\S_{\v m}}$
as a basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\hat\S_{\v m}\equiv \hat S_+^{m_+} \hat S_\z^{m_\z} \hat S_-^{m_-}$,
we can expand any collective spin operator $\hat\O$ in the form
\begin{align}
  \hat\O = \sum_{\v m} \O_{\v m} \hat\S_{\v m}
\end{align}
with scalar coefficients $\O_{\v m}\in\mathbb{C}$.  If $\hat\O$ is
self-adjoint, for example, then $\O_{\v m}^*=\O_{\v m^*}$ with
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The corresponding Heisenberg operator
is then $\hat\O\p{t}=\sum_{\v m}\O_{\v m}\p{t}\hat\S_{\v m}$, with
time-dependent coefficients $\O_{\v n}\p{t}$ for time-independent
(Schr\"odinger) operators $\hat\S_{\v m}$.  The time evolution of
Heisenberg operators is determined by
\begin{align}
  \f{d}{dt} \hat\O\p{t}
  = \check T \hat\O\p{t}
  = \sum_{\v m, \v n} \hat\S_{\v m} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:EOM}
\end{align}
for a time derivative (super-)operator $\check T$ with matrix elements
$T_{\v m\v n}\in\mathbb{C}$ defined by
\begin{align}
  \check T \hat\S_{\v n} \equiv i \sp{\hat H, \hat\S_{\v n}}_-
  + \sum_\A \gamma_\A \check\D\p{\A} \hat\S_{\v n}
  = \sum_{\v m} \hat\S_{\v m} T_{\v m\v n},
  \label{eq:time_deriv}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\hat H$ is the collective spin
Hamiltonian; $\A$ is a set of jump operators with a corresponding
decoherence rate $\gamma_\A$; and $\check\D$ is a Heisenberg-picture
dissipator, or Lindblad superoperator, defined by
\begin{align}
  \check\D\p{\A} \hat \O
  \equiv \sum_{\hat A\in\A} \p{\hat A^\dag \hat \O \hat A
    - \f12\sp{\hat A^\dag \hat A, \hat \O}_+}.
\end{align}
Decoherence via uncorrelated decay of individual spins, for example,
would be described by the set of jump operators
$\A_-=\set{\hat s_-^{(j)}:j=1,2,\cdots,N}$.  The commutator in
Eq.~\eqref{eq:time_deriv} can be computed by expanding the product
$\hat\S_{\v\ell}\hat\S_{\v m} = \sum_{\v n} f_{\v\ell\v m\v
  n}\hat\S_{\v n}$ with structure constants
$f_{\v\ell\v m\v n}\in\mathbb{R}$ that we work out in Appendices
\ref{sec:identities}--\ref{sec:prod_general}, and the effects of
decoherence from jump operators (i.e.~elements of $\A$) of the form
$\hat g^{(j)} = \sum_\a g_\a \hat s_\a^{(j)}$ and
$\hat G = \sum_\a G_\a \hat S_\a$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be some of the main technical
contributions of this work, with potential applications beyond the
short-time simulation method presented here.  These ingredients are
sufficient to compute matrix elements $T_{\v m\v n}$ of the time
derivative operator $\check T$ in Eq.~\eqref{eq:time_deriv} in most
cases of practical interest.

@@

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \hat\O\p{t}
  = e^{t\check T} \hat \O\p{0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \hat\O\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Unlike the truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
which operators are relevant for computing the expectation value
$\bk{\hat\O\p{t}}$ to a fixed order $M$.  Note that using the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can cut both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}_{t=0}$ and the
number qof matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the expansion in Eq.~\eqref{eq:time_series} implies for any given time
$t$, there exists a truncation order $M_t$ for which the truncation
error in Eq.~\eqref{eq:TST} is negligibly small.  In practice, only a
maximal truncation order $M_{\t{max}}$ is accessible with limited
computational resources, such that the TST expansion in
Eq.~\eqref{eq:TST} only allows for computation of the expectation
value $\bk{\hat\O\p{t}}$ to a maximal time $t_{M_{\t{max}}}$.  As we
will see in the following section, $M_{\t{max}}=35$ will suffice for
the computation of collective spin correlators up to times that are
relevant e.g.~for spin squeezing protocols.

@@

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  The general type of
    decoherence (namely spin decay, excitation, or dephasing) also
    leaves a signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results
    computed using the TST expansion in Eq.~\eqref{eq:TST} with
    $M=20$.}
  \label{fig:two_time}
\end{figure*}

@@

\section{Product of arbitrary ordered collective spin operators}
\label{sec:prod_general}

@@

@@

Here, we discuss the appearance of two-time correlation functions in
the short-time linear response of Heisenberg operators to
perturbations of a Hamiltonian.  Consider an initial Hamiltonian $H$
perturbed by an operator $V$ with $\norm{V}\ll\norm{H}$, where
$\norm{\O}$ denotes the operator norm of $\O$, such that the net
Hamiltonian is $\widetilde H=H+V$.  We denote the generator of
Heisenberg time evolution under the perturbed (unperturbed)
Hamiltonian by $\widetilde T$ ($T$).  These generators are related by
\begin{align}
  \widetilde T = T + i\check V
\end{align}
where $\check V$ is a superoperator whose action on operators $\O$ is
defined by
\begin{align}
  \check V \O \equiv \sp{V,\O}_-.
\end{align}
Through quadratic order in the time $t$ and linear order in the
perturbation $\check V$, we can say that
\begin{align}
  e^{t\widetilde T}
  \approx \f12\sp{e^{tT}, e^{it\check V}}_+
  \approx e^{tT} + \f12 it \sp{e^{t T}, \check V}_+.
\end{align}
Defining perturbed and unperturbed Heisenberg operators
$\widetilde\O\p{t}\equiv e^{t\widetilde T}\O$ and
$\O\p{t}\equiv e^{tT}\O$, we thus find that for sufficiently small
times $t$ and weak perturbations $V$,
\begin{align}
  \widetilde\O\p{t} - \O\p{t}
  = \p{e^{t\widetilde T} - e^{tT}} \O
  \approx \f12 i t \p{\sp{V,\O}_-\p{t} + \sp{V,\O\p{t}}_-}.
  \label{eq:response}
\end{align}
Two-time operators $V\O\p{t}$ and $\O\p{t} V$, in addition to the
Heisenberg operators $\p{V\O}\p{t}$ and $\p{\O V}\p{t}$, thus
determine the short-time linear response of operators $\O\p{t}$ to
perturbations $V$ of a Hamiltonian.

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\I}{\mathcal{I}}
\renewcommand{\O}{\mathcal{O}}
\renewcommand{\S}{\mathcal{S}}

\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}

@@

In this work, we present a new method to compute short-time dynamics
of collective spin systems with generic Markovian decoherence.  The
only restriction on decoherence is that, like the coherent collective
dynamics, it must act identically on all constituent particles.  This
method is based on a short-time expansion of exact solutions to the
equations of motion for Heisenberg operators.  Evaluating this
expansion requires knowing the structure constants of a collective
spin operator algebra; the calculation of these structure constants
(in Appendices \ref{sec:identities}--\ref{sec:general_product}) is one
of the main technical results of this work, which we hope will empower
both analytical and numerical studies of collective spin systems in
the future.  We benchmark our method against both exact and quantum
trajectory Monte Carlo computations of spin squeezing in accessible
parameter regimes, highlighting both advantages and limitations of our
short-time expansion.  Finally, we showcase applications of our method
with computations that are inaccessible to other numerical methods.

@@

In this section we provide the basic theory for our method to compute
expectation values of collective spin operators, deferring lengthy
derivations to the appendices.  We consider systems composed of $N$
distinct spin-1/2 particles.  Defining individual spin-1/2 operators
$\hat s_{\a=\x,\y,\z}\equiv\hat\sigma_\a/2$ and
$\hat s_\pm\equiv\hat s_\x\pm i\hat s_\y=\hat\sigma_\pm$ with Pauli
operators $\hat\sigma_\a$, we denote an operator that acts with
$\hat s_\a$ on the spin indexed by $j$ and trivially (i.e.~with the
identity $\hat\1$) on all other spins by $\hat s_\a^{(j)}$.  We then
define the collective spin operators
$\hat S_\a\equiv\sum_{j=1}^N\hat s_\a^{(j)}$ for
$\a\in\set{\x,\y,\z,+,-}$.  Identifying the set $\set{\hat\S_{\v m}}$
as a basis for all collective spin operators, with
$\v m\equiv\p{m_+,m_\z,m_-}\in\N_0^3$ and
$\hat\S_{\v m}\equiv \hat S_+^{m_+} \hat S_\z^{m_\z} \hat S_-^{m_-}$,
we can expand any collective spin operator $\hat\O$ in the form
\begin{align}
  \hat\O = \sum_{\v m} \O_{\v m} \hat\S_{\v m}
\end{align}
with scalar coefficients $\O_{\v m}\in\C$.  If $\hat\O$ is
self-adjoint, for example, then $\O_{\v m}^*=\O_{\v m^*}$ with
$\v m^*\equiv\p{m_-,m_\z,m_+}$.  The corresponding Heisenberg operator
is then $\hat\O\p{t}=\sum_{\v m}\O_{\v m}\p{t}\hat\S_{\v m}$, with
time-dependent coefficients $\O_{\v n}\p{t}$ for time-independent
(Schr\"odinger) operators $\hat\S_{\v m}$.  The time evolution of
Heisenberg operators is determined by
\begin{align}
  \f{d}{dt} \hat\O\p{t}
  = \check T \hat\O\p{t}
  = \sum_{\v m, \v n} \hat\S_{\v m} T_{\v m\v n} \O_{\v n}\p{t}
  \label{eq:EOM}
\end{align}
for a time derivative (super-)operator $\check T$ with matrix elements
$T_{\v m\v n}\in\C$ defined by
\begin{align}
  \check T \hat\S_{\v n} \equiv i \sp{\hat H, \hat\S_{\v n}}_-
  + \sum_\A \gamma_\A \check\D\p{\A} \hat\S_{\v n}
  = \sum_{\v m} \hat\S_{\v m} T_{\v m\v n},
  \label{eq:time_deriv}
\end{align}
where $\sp{X,Y}_\pm\equiv XY\pm YX$; $\hat H$ is the collective spin
Hamiltonian; $\A$ is a set of jump operators with a corresponding
decoherence rate $\gamma_\A$; and $\check\D$ is a Heisenberg-picture
dissipator, or Lindblad superoperator, defined by
\begin{align}
  \check\D\p{\A} \hat \O
  \equiv \sum_{\hat A\in\A} \p{\hat A^\dag \hat \O \hat A
    - \f12\sp{\hat A^\dag \hat A, \hat \O}_+}.
\end{align}
Decoherence via uncorrelated decay of individual spins, for example,
would be described by the set of jump operators
$\A_-\equiv\set{\hat s_-^{(j)}:j=1,2,\cdots,N}$.  The commutator in
Eq.~\eqref{eq:time_deriv} can be computed by expanding the product
$\hat\S_{\v\ell}\hat\S_{\v m} = \sum_{\v n} f_{\v\ell\v m\v
  n}\hat\S_{\v n}$ with structure constants
$f_{\v\ell\v m\v n}\in\mathbb{R}$ that we work out in Appendices
\ref{sec:identities}--\ref{sec:general_product}, and the effects of
decoherence from jump operators (i.e.~elements of $\A$) of the form
$\hat g^{(j)} = \sum_\a g_\a \hat s_\a^{(j)}$ and
$\hat G = \sum_\a G_\a \hat S_\a$ are worked out in Appendices
\ref{sec:sandwich_single}--\ref{sec:decoherence_collective}.  We
consider these calculations to be some of the main technical
contributions of this work, with potential applications beyond the
short-time simulation method presented here.  These ingredients are
sufficient to compute matrix elements $T_{\v m\v n}$ of the time
derivative operator $\check T$ in Eq.~\eqref{eq:time_deriv} in most
cases of practical interest.

@@

We can formally expand Heisenberg operators $\hat\O\p{t}$ in a Taylor
series about the time $t=0$ to write
\begin{align}
  \hat\O\p{t}
  = e^{t\check T} \hat \O\p{0}
  = \sum_{k\ge0} \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0},
  \label{eq:time_series}
\end{align}
where the matrix elements $T^k_{\v m\v n}$ of the $k$-th time
derivative operator $T^k$ are
\begin{align}
  T^0_{\v m\v n} &\equiv \delta_{\v m\v n}, \\
  T^1_{\v m\v n} &\equiv T_{\v m\v n}, \\
  T^{k>1}_{\v m\v n}
  &\equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n},
\end{align}
with $\delta_{\v m\v n}=1$ if $\v m=\v n$ and 0 otherwise.  For
sufficiently short times, we can truncate the series in
Eq.~\eqref{eq:time_series} by taking
\begin{align}
  \hat\O\p{t}
  \to \sum_{k=0}^M \f{t^k}{k!}
  \sum_{\v m, \v n} \hat\S_{\v m} T^k_{\v m\v n} \O_{\v n}\p{0}.
  \label{eq:TST}
\end{align}
We refer Eq.~\eqref{eq:TST} as the truncated short-time (TST)
expansion of Heisenberg operators.  Unlike the truncation in
Eq.~\eqref{eq:weight_truncation}, the nonzero matrix elements
$T^k_{\v m\v n}$ for $k=0,1,\cdots,M$ in Eq.~\eqref{eq:TST} tell us
which operators are relevant for computing the expectation value
$\bk{\hat\O\p{t}}$ to a fixed order $M$.  Note that using the relation
$\hat\S_{\v m}^\dag=\hat\S_{\v m^*}$, which by Hermitian conjugation
of Eq.~\eqref{eq:EOM} also implies that
$T_{\v m^*\v n^*}=T_{\v m\v n}^*$, can cut both the number of
initial-time expectation values $\bk{\hat\S_{\v m}}_{t=0}$ and the
number qof matrix elements $T_{\v m\v n}$ that we may need to
explicitly compute roughly in half.

In principle, the factorial suppression of terms at higher orders of
the expansion in Eq.~\eqref{eq:time_series} implies for any given time
$t$, there exists a truncation order $M_t$ for which the truncation
error in Eq.~\eqref{eq:TST} is negligibly small.  In practice, only a
maximal truncation order $M_{\t{max}}$ is accessible with limited
computational resources, such that the TST expansion in
Eq.~\eqref{eq:TST} only allows for computation of the expectation
value $\bk{\hat\O\p{t}}$ to a maximal time $t_{M_{\t{max}}}$.  As we
will see in the following section, $M_{\t{max}}=35$ will suffice for
the computation of collective spin correlators up to times that are
relevant e.g.~for spin squeezing protocols.  Appendix
\ref{sec:tutorial} provides a pedagogical tutorial for computing
correlators using the TST expansion.

@@

\begin{figure*}
  \centering
  \includegraphics{two_time.pdf}
  \caption{Magnitude and phase of the two-time connected correlator
    $C\p{t}\equiv\abs{C\p{t}}\exp\sp{i\phi\p{t}}$ defined in
    Eq.~\eqref{eq:two_time} for $N=10^4$ spins initially in the
    polarized state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
    evolving under the squeezing Hamiltonians in
    Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT}.  Results are shown for time
    evolution without decoherence (leftmost panel), as well as with
    decoherence at rates $\Gamma_\a$ or $\gamma_\a$, respectively
    corresponding to collective or single-spin decoherence through
    jump operators $\hat S_\a$ or $\hat s_\a^{(j)}$.  Collective
    decoherence rates ($\Gamma_\a$) were chosen to be comparable to
    the OAT squeezing strength $\chi$, while single-spin decoherence
    rates ($\gamma_\a$) were scaled by an additional factor of $S=N/2$
    in order to yield comparable effects on the correlator $C\p{t}$ at
    the times shown.  At short times, the correlations captured by
    $C\p{t}$ are essentially insensitive to collective dephasing
    ($\Gamma_\z$), while the magnitude $\abs{C\p{t}}$ is generally
    suppressed by other forms of decoherence.  Whether decoherence
    occurs through spin decay, excitation, or dephasing also leaves a
    signature on the phase $\phi\p{t}$ of $C\p{t}$.  Results computed
    using the TST expansion in Eq.~\eqref{eq:TST} with $M=20$.}
  \label{fig:two_time}
\end{figure*}

@@

\section{Product of arbitrary ordered collective spin operators}
\label{sec:general_product}

@@

\section{Computing correlators with the truncated short-time (TST)
  expansion}
\label{sec:tutorial}

Here, we provide a pedagogical tutorial for computing correlators
using the truncated short-time TST expansion.  For concreteness, we
nominally consider $N$ spins evolving under the one-axis twisting
(OAT) Hamiltonian
\begin{align}
  H_{\t{OAT}} = \chi S_\z^2,
\end{align}
additionally subject to spontaneous single-spin decay at rate
$\gamma_-$, with jump operators $\A_-=\set{s_-^{(j)}:j=1,2,\cdots,N}$.
The equation of motion for a Heisenberg operator
$S_+^\ell S_\z^m S_-^n$ is
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n}
  = i\chi\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-
  + \gamma_- \D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n},
  \label{eq:tutorial_EOM}
\end{align}
where we can use the results in appendices \ref{sec:general_product}
and \ref{sec:decoherence_single} respectively to evaluate the
commutator $\sp{S_\z^2, S_+^\ell S_\z^m S_-^n}_-$ and dissipator
$\D\p{\A_-} \p{S_+^\ell S_\z^m S_-^n}$ in \eqref{eq:tutorial_EOM} (see
\eqref{eq:general_product} and \eqref{eq:decay_diff} in particular),
giving us
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_\z^m S_-^n} =
  i\chi \p{\ell-n} S_+^\ell \p{\ell+n+2S_\z} S_\z^m S_-^n
  + \gamma_- S_+^\ell \sp{\p{S+S_\z}\p{-1+S_\z}^m
    - \p{S+\f{\ell+n}{2}+S_\z} S_\z^m} S_-^n.
  \label{eq:tutorial_EOM_expanded}
\end{align}
In practice, we do not want to keep track of such an expansion by
hand, especially in the case of e.g.~the two-axis twisting (TAT) and
twist-and-turn (TNT) models with more general types of decoherence,
for which the analogue of \eqref{eq:tutorial_EOM_expanded} may take
several lines just to write out in full.  Defining the operators
$\S_{\v m}\equiv S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ with
$\v m\equiv\p{m_+,m_\z,m_-}$ for shorthand, we note that the vector
space spanned by $\set{\S_{\v m}}$ is closed under time evolution.  We
therefore expand
\begin{align}
  \f{d}{dt} \S_{\v n} \equiv T \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n},
  \label{eq:tutorial_EOM_general}
\end{align}
where $T$ is a super-operator that generates time evolution for
Heisenberg operators.  In the present example, the matrix elements
$T_{\v m\v n}\in\C$ of $T$ are defined by
\eqref{eq:tutorial_EOM_expanded} and \eqref{eq:tutorial_EOM_general}.
For any Hamiltonian $H$ with decoherence characterized by sets of jump
operators $\A$ and decoherence rates $\gamma_A$, the matrix elements
$T_{\v m\v n}$ are more generally defined by
\begin{align}
  T \S_{\v n}
  = i\sp{H,\S_{\v n}}_- + \sum_\A \gamma_\A \D\p{\A} \S_{\v n}
  = \sum_{\v m} \S_{\v m} T_{\v m\v n}.
\end{align}
The results in Appendices \ref{sec:general_product},
\ref{sec:decoherence_single}, and \ref{sec:decoherence_collective} can
be used to write model-agnostic codes that compute matrix elements
$T_{\v m\v n}$, taking a particular Hamiltonian $H$ and decoherence
processes $\set{\p{\A,\gamma_\A}}$ as inputs.

In order to compute a quantity such as spin squeezing, we need to
compute correlators of the form $\bk{\S_{\v n}\p{t}}$, where
$\S_{\v n}\p{t}$ is the time-evolved Heisenberg operator corresponding
to $\S_{\v n}$, with $\S_{\v n}\p{0}\equiv\S_{\v n}$.  The order-$M$
truncated short-time (TST) expansion takes
\begin{align}
  \bk{\S_{\v n}\p{t}}
  \to \sum_{k=0}^M \f{t^k}{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
  \label{eq:tutorial_TST}
\end{align}
where $T^k_{\v m\v n}$ are matrix elements of the $k$-th time
derivative operator, given by
\begin{align}
  T^0_{\v m\v n} \equiv
    \begin{cases}
      1 & \v m = \v n, \\
      0 & \t{otherwise}
    \end{cases},
  &&
  T^1_{\v m\v n} \equiv T_{\v m\v n},
  &&
  T^{k>1}_{\v m\v n}
  \equiv \sum_{\v p_1,\v p_2,\cdots,\v p_{k-1}}
  T_{\v m\v p_{k-1}} \cdots T_{\v p_3\v p_2}
  T_{\v p_2\v p_1} T_{\v p_1\v n}.
\end{align}
Matrix elements $T^k_{\v m\v n}$ and initial-time expectation values
$\bk{\S_{\v m}}$ are thus computed on an as-needed basis for any
particular correlator $\bk{\S_{\v n}\p{t}}$ of interest, and combined
according to \eqref{eq:tutorial_TST}.  Note that initial-time
expectation values $\bk{\S_{\v m}}$ are an input to the TST expansion,
and need to be computed separately for any particular initial state of
interest; expectation values with respect to spin-polarized (Gaussian)
states are provided in Appendix \ref{sec:initial_conditions}.  In
practice, we further collect terms in \eqref{eq:tutorial_TST} to write
\begin{align}
  \bk{\S_{\v n}\p{t}} \to \sum_{k=0}^M c_{\v n k} t^k,
  &&
  c_{\v n k} \equiv \f1{k!} \sum_{\v m} \bk{\S_{\v m}} T^k_{\v m\v n},
\end{align}
where $c_{\v n k}$ are time-independent coefficients for the
polynomial time expansion of $\bk{\S_{\v n}\p{t}}$.  After computing
the coefficients $c_{\v n k}$, there is only negligible computational
overhead to compute the correlator $\bk{\S_{\v n}\p{t}}$ for any time
$t$ up to the breakdown time $t_M$ of an order-$M$ TST expansion.

@@

Here, we discuss the appearance of two-time correlation functions in
the short-time linear response of correlators to perturbations of a
Hamiltonian.  Consider an initial Hamiltonian $H$ perturbed by an
operator $V$ with $\norm{V}\ll\norm{H}$, where $\norm{\O}$ denotes the
operator norm of $\O$, such that the net Hamiltonian is
$\widetilde H=H+V$.  We denote the generator of Heisenberg time
evolution under the perturbed (unperturbed) Hamiltonian by
$\widetilde T$ ($T$).  These generators are related by
\begin{align}
  \widetilde T = T + i\check V
\end{align}
where $\check V$ is a superoperator whose action on operators $\O$ is
defined by
\begin{align}
  \check V \O \equiv \sp{V,\O}_-.
\end{align}
Through quadratic order in the time $t$ and linear order in the
perturbation $\check V$, we can say that
\begin{align}
  e^{t\widetilde T}
  \approx \f12\sp{e^{tT}, e^{it\check V}}_+
  \approx e^{tT} + \f12 it \sp{e^{t T}, \check V}_+.
\end{align}
Defining perturbed and unperturbed Heisenberg operators
$\widetilde\O\p{t}\equiv e^{t\widetilde T}\O$ and
$\O\p{t}\equiv e^{tT}\O$, we thus find that for sufficiently small
times $t$ and weak perturbations $V$,
\begin{align}
  \widetilde\O\p{t} - \O\p{t}
  = \p{e^{t\widetilde T} - e^{tT}} \O
  \approx \f12 i t \p{\sp{V,\O}_-\p{t} + \sp{V,\O\p{t}}_-}.
  \label{eq:response}
\end{align}
Two-time operators $V\O\p{t}$ and $\O\p{t} V$, in addition to the
Heisenberg operators $\p{V\O}\p{t}$ and $\p{\O V}\p{t}$, thus
determine the short-time linear response of operators $\O\p{t}$ to
perturbations $V$ of a Hamiltonian.
